---
title: "GBIF Crayfish Data Workflow"
format: html
editor: visual
execute:
  echo: true
  warning: false
---

## Overview

This workflow demonstrates how to:

-   [Retrieve specimen occurrence data from GBIF](#sec-gbif)
-   Filter records to a focal taxon
-   Convert tabular data into spatial objects
-   Generate both interactive and static maps

This example uses crayfish (Family: Cambaridae) in Alabama.

## Load Packages & Import Functions

```{r}
#| warning: false

# Packages
if (!require("pacman")) install.packages("pacman"); library(pacman)
p_load(dplyr, sf, ggplot2, leaflet, ggplot2,
       install = TRUE, update = FALSE)

# Functions
source("R/gbif_utils.R")
source("R/spatial_utils.R")
source("R/viz_utils.R")
source("R/gadm_utils.R")
```

## Retrieve GBIF Data {#sec-gbif}

::: callout-note
GBIF records could include:

-   Missing coordinates — *these records will not be captured in this workflow*
-   Synonyms vs accepted names
-   Multiple identifications per occurrence\*
:::

1.  Define what you'll be searching for.

In this case, I want to search for all of our crayfish here in the family Cambaridae.

```{r}
taxon_key <- get_taxon_key("Lacunicambarus")
```

2.  Define the geographic area.

I want to find everything in Alabama, so I can use the function `lookup_state_gadm()`. This is better than providing a state name because not all data providers fill that information in, but the gadm is assigned by GBIF as long as the record has point coordinates.

```{r}
state_gadm <- lookup_state_gadm(state_name = "alabama")
```

```{r}
gbif_raw <- get_gbif_occurrences(
  taxon_key = taxon_key,
  state_gadm = state_gadm,
  # collection_code = "Invertebrates",
  # publishing_org = "34cc6c50-812c-11de-86fe-b8a03c50a862"
)
```

```{r}
occ_df <- extract_occurrence_data(gbif_raw)
```

## Explore & Filter Taxa

```{r}
sort(unique(occ_df$acceptedScientificName))
```

This helps verify the taxonomic composition of returned records.

```{r}
focal_df <- filter_by_level(occ_df, tax_level = "genus", search_name = "Lacunicambarus")
```

## Convert to Spatial Data

```{r}
state_sf <- get_state_boundary("alabama")

points_sf <- to_sf_points(focal_df)

points_clipped <- clip_to_boundary(input_sf = points_sf, bounding_sf = state_sf)
```

### Static Map

```{r}
plot_occurrences_map(state_sf, points_clipped)
```

### Interactive Map

```{r}
leaflet_map <- make_leaflet_map(focal_df)

leaflet_map
```

You may choose to save the output...

```{r}
#| eval: false
# htmlwidgets::saveWidget(leaflet_map, "crayfish_map.html")
```

### Additional Layers

#### County

```{r}
counties <- st_as_sf(maps::map("county", 'alabama', fill = TRUE, plot = FALSE)) |>
  st_set_crs(4326)
```

#### HUC-Levels

```{r}
library(nhdplusTools)

# Query and download HUC8 watersheds overlapping Alabama's bounding box
huc8 <- get_huc(AOI = state_sf$geom, type = "huc08") # Fetches HUC8s within the AOI

# Ensure the coordinate reference systems (CRS) match for clipping
huc8 <- st_transform(huc8, st_crs(state_sf))

# Optional: Intersect/clip the watersheds strictly to the state border
# -- Remove this step if you want to keep the full watersheds extending past state lines
huc8_clipped <- st_intersection(huc8, state_sf)

# Optional: Extract HUC6 and create grouped colors
# -- 'id' should contiain the 8-digit string...
library(dplyr)
library(scico)

huc8_clipped_styled <- huc8_clipped |>
  mutate(
    huc6 = substr(id, 1, 6) # Extract first 6 digits to get parent HUC6
  ) |>
  arrange(huc6, id) # Group sequentially for color palette

# -- Pull HUC6 metadata
huc6_metadata <- get_huc(AOI = state_sf, type = "huc06") |>
  st_drop_geometry() |>
  select(huc6 = id, huc6_name = name)

# -- Join to HUC8 layer
huc8_clipped_styled <- huc8_clipped |>
  mutate(huc6 = substr(id, 1, 6)) |>
  left_join(huc6_metadata, by = "huc6") |> # Append text names
  mutate(legend_label = paste0(huc6, " - ", huc6_name)) |>
  arrange(huc6) # Keeps colors sorted sequentially

# -- Combine HUC8 geoms into HUC6 for a border at HUC6 level...
huc_6_boundaries <- huc8_clipped_styled |>
  group_by(legend_label) |>
  summarize(geometry = st_union(geometry), .groups = "drop")
```

### Final Map (Interactive)

[More information about the R package {leaflet}](https://rstudio.github.io/leaflet/)

```{r}
library(htmltools)
# Format for popup hyperlinks to GBIF record
focal_df$popup_links <- lapply(1:nrow(focal_df), function(i) {
  # Combine the base URL with the specific occurrence key
  full_url <- paste0("https://www.gbif.org/occurrence/", focal_df$key[i])
  
  HTML(sprintf(
    "<b>Species:</b> <i>%s</i><br>
      <b>Cat. Number:</b> %s <br>
      <b>Location:</b> %s <br>
      <b>Date:</b> %s <br>
      <br>
     <a href='%s' target='_blank' rel='noopener noreferrer'>View on GBIF &rarr;</a>",

    if (is.na(focal_df$species[i])) {focal_df$genus[i]} else {focal_df$species[i]},
    htmlEscape(paste(focal_df$institutionCode[i], focal_df$collectionCode[i], focal_df$catalogNumber[i], sep = ":")),
    focal_df$locality[i],
    focal_df$eventDate[i],
    full_url
  ))
})

```

Since there's not a field that gives the full taxon without the authority, create a "clean" species label to group consistently.

```{r}
focal_df$taxon_label <- ifelse(
  is.na(focal_df$species),
  focal_df$genus,
  focal_df$species
)
```

Split the data by `taxon_label`. This is important if you want to color the markers differently.

```{r}
species_list <- split(focal_df, focal_df$taxon_label)
```

In this, I'm using `addCircleMarker()`. It allows for easier color manipulation, if that's wanted.

```{r}
# How many colors are needed? 
length(unique(focal_df$scientificName))

library(ggokabeito)
library(impressionist.colors)
pal <- colorFactor(
  # palette = scico(5, palette = "bilbao"),
  palette = impressionist.colors::get.color(8, 2, 1:16),
  domain = focal_df$taxon_label
)

# For including colors in the legend.
legend_colors <- pal(names(species_list))

# Species labels...
species_labels <- sprintf(
  "<span style = 'color:%s;'>■</span> <i>%s</i>",
  pal(names(species_list)),
  names(species_list)
)
```

You may choose to have the data represented by icons, which you can customize. When you use `makeAwesomeIcon()`, the `markerColor` is restricted to certain named values.

```{r}
#| label: make-icon
awesome <- makeAwesomeIcon(
  icon = "info",
  library = "fa",
  markerColor = "black", # Color of background pin
  # markerColor = pal(focal_df$scientificName),
  iconColor = pal(focal_df$scientificName)
  # iconColor = "black" # Color of the icon itself
)
```

::: callout-note
The icons available, at least for `library = "fa"`, come from an old fontawesome version. It seems like the full list of options can be found on the [Font Awesome 4.7.0 Cheatsheet](https://fontawesome.com/v4/cheatsheet/).
:::

```{r}
#leaflet_map %>%
library(leaflet.extras)

  ## --- Build Base Map --- ##
leaflet_map <-  leaflet(focal_df) |>
  leaflet::addProviderTiles(leaflet::providers$Esri.NatGeoWorldMap, group = "NatGeo World (default") |>
  leaflet::addTiles(group = "Open Streets Map")
    # leaflet::addAwesomeMarkers(
    #   lng = ~decimalLongitude,
    #   lat = ~decimalLatitude,
    #   group = "Points",
    #   icon = awesome,
    #   popup = ~popup_links,
    #   clusterOptions = leaflet::markerClusterOptions(
    #     maxClusterRadius = 50,
    #   )
    # ) %>%
  
## -- Define Panes -- ##
leaflet_map <- leaflet_map %>%
  addMapPane("polygons", zIndex = 400) |>
  addMapPane("points", zIndex = 500)

## --- Add Polygons --- ###
leaflet_map <- leaflet_map |>
  ## --- Add Overlay Polygons --- ##
  addPolygons(data = state_sf,
              group = "State Boundary",
              color = "black",
              weight = 5,
              opacity = 1
              ) |>
  addPolygons(data = counties, 
              group = "Counties",
              options = pathOptions(pane = "polygons"),
              fillColor = "transparent",
              color = "black",
              label = ~unname(stringr::str_to_title(stringr::str_split_i(counties$ID, ",",2))),
              highlightOptions = highlightOptions(
                color = "white",
                fillColor = "grey",
                fillOpacity = 0.2,
                weight = 5
              )) |>
  addPolygons(data = huc8_clipped_styled, 
              group = "HUC8",
              options = pathOptions(pane = "polygons"),
              fillColor = "transparent",
              color = "blue",
              label = ~name,
              highlightOptions = highlightOptions(
                color = "grey20",
                fillOpacity = 0.2,
                fillColor = "blue",
                weight = 5
              )) |>
  addPolygons(data = huc_6_boundaries, 
              group = "HUC6",
              options = pathOptions(pane = "polygons"),
              fillColor = unname(scico(nrow(huc_6_boundaries), palette = "batlowK")), 
              fillOpacity = 0.8, 
              stroke = FALSE,
              label = ~unname(stringr::str_split_i(huc_6_boundaries$legend_label, "- ", 2)),
              highlightOptions = highlightOptions(
                color = "grey20",
                fillOpacity = 0.2,
                weight = 5
              ))
  
  ## ----- Add Occurrence Points ----- ##
  for (i in names(species_list)) {
  df <- species_list[[i]]
  
  leaflet_map <- leaflet_map |>
    addCircleMarkers(
      data = df,
      lng = ~decimalLongitude,
      lat = ~decimalLatitude,
      group = i,
      options = pathOptions(pane = "points"),
      color = "white",
      weight = 1.75,
      opacity = 1, 
      fillColor = ~unname(pal(i)),
      fillOpacity = 1,
      radius = 7,
      popup = ~popup_links,
      clusterOptions = markerClusterOptions(
        maxClusterRadius = 150,
        spiderfyOnMaxZoom = FALSE,
        zoomToBoundsOnClick = TRUE,
        disableClusteringAtZoom = 7,
        polygonOptions = list(
          fillColor = "white",
          color = "white",
          weight = 3,
          opacity = 0.8,
          fillOpacity = 0.7
        ) 
      )
    )
}

  ## --- Add Grouped & Collapsible Layer Control --- ##
  leaflet_map <- leaflet_map |>
  # Which layers should be hidden by default
  hideGroup(c("Counties", "HUC8")) |>
  addGroupedLayersControl(
    baseGroups = c("NatGeo World (default)", "Open Streets Map"),
    overlayGroups = list(
      "Species" = setNames(names(species_list), species_labels),
       "Boundaries" = c("Counties", "HUC8", "HUC6")
    ),
    position = "topright",
    options = groupedLayersControlOptions(
      collapsed = FALSE,
      # groupsCollapsed = FALSE,
      groupsCollapsable = FALSE,
      sortGroups = TRUE,
      exclusiveGroups = c("Boundaries")  # allows multiple selections
    )
  ) %>%
      addControl("All GBIF Occurrences", position = "bottomleft") 
## -- Add Legend -- ##
# leaflet_map %>%
#     addLegend(
#       position = "bottomright",
#       colors = legend_colors,
#       labels = sprintf("<i>%s</i>", names(species_list)),
#       title = "Species",
#       opacity = 1
#     )
leaflet_map
```

Then if we want to add our "home bases" of Auburn and Troy...

```{r}
library(nominatimlite) # queries OpenStreetMap's Nominatim API, geocoding

# Look up the place by name
auburn <- geo_lite_sf(
  address = "Auburn University, Alabama",
  points_only = TRUE 
)

troy <- geo_lite_sf(
  address = "Troy University, Alabama",
  points_only = TRUE
)

poi <- rbind(auburn, troy)
```

```{r}
awesome <- makeAwesomeIcon(
  icon = "university",
  library = "fa",
  markerColor = "black", # Color of background pin
  iconColor = "#FFF"
)
```

```{r}
leaflet_map <- leaflet_map %>%
  addAwesomeMarkers(
    data = poi,
    icon = awesome,
    popup = ~paste0("<b>",stringr::str_split_i(query, ",", 1),"</b>")
      )

```

[How to Save a Leaflet Map as a html widget in R](https://www.geeksforgeeks.org/r-language/how-to-save-a-leaflet-map-as-a-html-widget-in-r/)

Also, the R package [{leaflegend}](https://leaflegend.delveds.com) looks very promising!

```{r}
library(htmlwidgets)
saveWidget(leaflet_map, file = "concept-demos/saved-output/GBIF_Lacunicambarus_Alabama.html", selfcontained = TRUE)
```

# 
