---
title: "GBIF Crayfish Data Workflow"
format: html
editor: visual
execute:
  echo: true
  warning: false
---

## Overview

This workflow demonstrates how to:

- [Retrieve specimen occurrence data from GBIF](#sec-gbif)
- Filter records to a focal taxon
- Convert tabular data into spatial objects
- Generate both interactive and static maps

This example uses crayfish (Family: Cambaridae) in Alabama.

## Load Packages & Import Functions
```{r}
#| warning: false

# Packages
if (!require("pacman")) install.packages("pacman"); library(pacman)
p_load(dplyr, sf, ggplot2, leaflet, ggplot2,
       install = TRUE, update = FALSE)

# Functions
source("R/gbif_utils.R")
source("R/spatial_utils.R")
source("R/viz_utils.R")
source("R/gadm_utils.R")
```

## Retrieve GBIF Data {#sec-gbif}

::: {.callout-note}
GBIF records could include:

* Missing coordinates — *these records will not be captured in this workflow*
* Synonyms vs accepted names
* Multiple identifications per occurrence*

:::

1. Define what you'll be searching for.

In this case, I want to search for all of our crayfish here in the family Cambaridae.

```{r}
taxon_key <- get_taxon_key("Cambaridae")
```

2. Define the geographic area.

I want to find everything in Alabama, so I can use the function `lookup_state_gadm()`. This is better than providing a state name because not all data providers fill that information in, but the gadm is assigned by GBIF as long as the record has point coordinates. 

```{r}
state_gadm <- lookup_state_gadm(state_name = "alabama")
```

```{r}
gbif_raw <- get_gbif_occurrences(
  taxon_key = taxon_key,
  state_gadm = state_gadm,
  collection_code = "Invertebrates",
  publishing_org = "34cc6c50-812c-11de-86fe-b8a03c50a862"
)
```

```{r}
occ_df <- extract_occurrence_data(gbif_raw)
```

## Explore & Filter Taxa

```{r}
sort(unique(occ_df$acceptedScientificName))
```

This helps verify the taxonomic composition of returned records.

```{r}
focal_df <- filter_by_level(occ_df, tax_level = "genus", search_name = "Lacunicambarus")
```

## Convert to Spatial Data

```{r}
state_sf <- get_state_boundary("alabama")

points_sf <- to_sf_points(focal_df)

points_clipped <- clip_to_boundary(input_sf = points_sf, bounding_sf = state_sf)
```

### Static Map
```{r}
plot_occurrences_map(state_sf, points_clipped)
```

### Interactive Map
```{r}
leaflet_map <- make_leaflet_map(focal_df)

leaflet_map
```

You may choose to save the output...

```{r}
#| eval: false
htmlwidgets::saveWidget(leaflet_map, "crayfish_map.html")
```

### Additional Layers 

#### County

```{r}
counties <- st_as_sf(maps::map("county", 'alabama', fill = TRUE, plot = FALSE)) |>
  st_set_crs(4326)
```


#### HUC-Levels

```{r}
library(nhdplusTools)

# Query and download HUC8 watersheds overlapping Alabama's bounding box
huc8 <- get_huc(AOI = state_sf$geom, type = "huc08") # Fetches HUC8s within the AOI

# Ensure the coordinate reference systems (CRS) match for clipping
huc8 <- st_transform(huc8, st_crs(state_sf))

# Optional: Intersect/clip the watersheds strictly to the state border
# -- Remove this step if you want to keep the full watersheds extending past state lines
huc8_clipped <- st_intersection(huc8, state_sf)

# Optional: Extract HUC6 and create grouped colors
# -- 'id' should contiain the 8-digit string...
library(dplyr)
library(scico)

huc8_clipped_styled <- huc8_clipped |>
  mutate(
    huc6 = substr(id, 1, 6) # Extract first 6 digits to get parent HUC6
  ) |>
  arrange(huc6, id) # Group sequentially for color palette

# -- Pull HUC6 metadata
huc6_metadata <- get_huc(AOI = state_sf, type = "huc06") |>
  st_drop_geometry() |>
  select(huc6 = id, huc6_name = name)

# -- Join to HUC8 layer
huc8_clipped_styled <- huc8_clipped |>
  mutate(huc6 = substr(id, 1, 6)) |>
  left_join(huc6_metadata, by = "huc6") |> # Append text names
  mutate(legend_label = paste0(huc6, " - ", huc6_name)) |>
  arrange(huc6) # Keeps colors sorted sequentially

# -- Combine HUC8 geoms into HUC6 for a border at HUC6 level...
huc_6_boundaries <- huc8_clipped_styled |>
  group_by(legend_label) |>
  summarize(geometry = st_union(geometry), .groups = "drop")
```


### Final Map (Interactive)
```{r}
# Format for popup hyperlinks to GBIF record
focal_df$popup_links <- lapply(1:nrow(focal_df), function(i) {
  # Combine the base URL with the specific occurrence key
  full_url <- paste0("https://www.gbif.org/occurrence/", focal_df$key[i])
  
  HTML(sprintf(
    "<b>Species:</b> <i>%s</i><br>
      <b>Cat. Number:</b> %s <br>
      <b>Location:</b> %s <br>
      <br>
     <a href='%s' target='_blank' rel='noopener noreferrer'>View on GBIF &rarr;</a>",

    if (is.na(focal_df$species[i])) {focal_df$genus[i]} else {focal_df$species[i]},
    htmlEscape(paste(focal_df$institutionCode[i], focal_df$collectionCode[i], focal_df$catalogNumber[i], sep = ":")),
    focal_df$locality[i],
    full_url
  ))
})

#leaflet_map %>%

leaflet_map <-  leaflet(focal_df) |>
    leaflet::addProviderTiles(leaflet::providers$Esri.NatGeoWorldMap) |>
    leaflet::addMarkers(
      lng = ~decimalLongitude,
      lat = ~decimalLatitude,
      group = "Points",
      popup = ~popup_links,
      clusterOptions = leaflet::markerClusterOptions()
    ) %>%
  addPolygons(data = state_sf,
              group = "State Boundary",
              color = "black",
              weight = 5,
              opacity = 1) %>%
  addPolygons(data = counties, 
              group = "Counties",
              fillColor = NA,
              color = "black",
              highlightOptions = highlightOptions(
                color = "white",
                fillColor = "grey",
                fillOpacity = 0.2,
                weight = 5
              )) %>%
  addPolygons(data = huc8_clipped_styled, 
              group = "HUC8",
              fillColor = NA,
              color = "blue",
              highlightOptions = highlightOptions(
                color = "grey20",
                fillOpacity = 0.2,
                fillColor = "blue",
                weight = 5
              )
              ) %>%
  addPolygons(data = huc_6_boundaries, 
              group = "HUC6", 
              fillColor = scico(nrow(huc_6_boundaries), palette = "batlowK"), 
              fillOpacity = 0.8, 
              stroke = FALSE,
              highlightOptions = highlightOptions(
                color = "grey20",
                fillOpacity = 0.2,
                weight = 5
              )) %>%
  # Layer Control
    addLayersControl(
    overlayGroups = c("Counties", "HUC8", "HUC6", "Points"),
    options = layersControlOptions(
      collapsed = TRUE
      )
  ) %>%
  # Which layers should be hidden
  hideGroup(c("Counties", "HUC8"))

leaflet_map
```

