---
title: "Trusting iNat Observations"
author: "BioCuration Lab"
format: html
---

```{r}
#| warning: false
if (!require("pacman")) install.packages("pacman"); library(pacman)
p_load(rinat, dplyr, stringr, tibble, ggplot2, sf, maps, patchwork, scico, gt, 
       httr2, purrr,
       tidyr,
       install = TRUE, update = FALSE)
```

The R package [`{rinat}`](https://github.com/ropensci/rinat) was made for interacting directly with the [iNaturalist API](https://api.inaturalist.org/v1/docs/). 

:::{.callout-tip}
If you run into any limitations, you could likely make a custom solution to the iNat API using {httr2}(https://httr2.r-lib.org).
:::

A useful vignette to get started...
```{r}
#| eval: false
vignette("rinat-intro", package = "rinat")
```

The quickest way is to search with `get_inat_obs()`.

You can refine the search using parameters.
 - `place_id`
 - `quality`
 - `taxon_name` - searches for it and all its descendants
 
 I'm not sure if there's a better way to do it, but I know that the iNat website has additive URLs. If I go to the normal *Explore* page, type in `Alabama`, hit search, and check the resulting URL... it looks like this: [https://www.inaturalist.org/observations?place_id=19](https://www.inaturalist.org/observations?place_id=19). Thus, I assume that the place_id for Alabama is 19.

```{r}
#| eval: true
#| output: false
obs <- get_inat_obs(
  taxon_name = "Orconectes australis",
  place_id = 19,
  quality = "research" # Only return research-grade observations
)
```

The `obs` object returns a ton of information...

```{r}
names(obs)
```

This information includes user attributes. If you have a list of people you trust to make the IDs, you can easily limit the `obs` object by any of these user attributes.

```{r}
obs %>%
  select(contains("user")) %>%
  names()
```

I'd recommend using either `user_login` or `user_name`. `user_id` may be a safer choice in the long run because it'll never change, but you'd need to pull it from iNat --- that could be really annoying!

## Establishing Your "Trusted" Users

Let's say you have a list of "trusted" users stored in a spreadsheet. For this example, I'm going to use the iNat IDs from these people providing their `user_login`...

1.  Import the trusted users

    ```{r}
    # trusted_users <- read.csv(...)
    
    trusted_users <- tibble(
      # Directly corresponds to iNaturalist field
      user_login = c("hydrophilus", "niemillernature", "kingjenna"),
      
      # Does NOT correspond to iNaturalist field
      name = c("Eric Maxwell", "Matt Niemiller", "Jenna King"),
      
      # Does NOT correspond to iNaturalist field
      affiliation = c(NA, "University of Alabama, Huntsville", "USFWS")
    )
    ```

    ```{r}
    trusted_users
    ```

2.  **Filter the rows in `obs` that match the `user_login` from the `trusted_users` data.frame.**

    ```{r}
    #| label: create-filtered-obs
    filtered_obs <- obs %>%
      filter(user_login %in% trusted_users$user_login)
    ```

    Do a small sanity check to make sure your filter actually worked...

    ```{r}
    # All observations...
    nrow(obs)
    
    # Observations only by trusted users...
    nrow(filtered_obs)
    ```

:::{.callout-important}
Use `filtered_obs` to pull any pertinent information you want for downstream analyses. 
:::

## Finding observations identified by trusted users...

This is going to be the "same" query as what I did above with rinat, but since I want to know more about who identified them, I need to do an API request with `{httr2}`. See the full list of potential parameters in the [iNaturalist API v2 documentation](https://api.inaturalist.org/v2/docs/#/Observations/get_observations).

```{r}
# 1. Prepare the request
req <- request("https://api.inaturalist.org/v2/observations") |>
  req_url_query(
    taxon_name = "Orconectes australis",
    place_id = 19,
    quality_grade = "research", # Only return research grade observations
    # What fields should be returned?
    fields = "all"
  )

# 2. Execute the request
resp <- req_perform(req)
data <- resp_body_json(resp, simplifyVector = TRUE)

# 3. Prep results to use
obs_api <- data$results
```

```{r}
names(obs_api)
```

```{r}
#| eval: false
# Exploring where stuff is at. This isn't super important to an end user.
class(obs_api$identifications)
class(obs_api$identifications[[1]]$user$login)
names(obs_api$identifications[[1]])

obs_api$taxon$name
obs_api$identifications[[1]]$taxon$name
obs_api$identifications[[1]]$user$login
```

 1. **Reshape identificaiton data**
```{r}
obs_tbl <- tibble(
  obs_id = obs_api$id, # observationID (not identificationID)
  identifications = obs_api$identifications
) %>%
  mutate(
    # Get all user logins associated with an observation
    logins = purrr::map(identifications, ~ .x$user$login)
    ) %>%
  # Expand so each row = one user login per observation
  unnest(logins) %>%
  rename(user_login = logins)
```

**What this does:**
* `obs_api$identifications` is a list where each element contains multiple IDs per observation.
* `map()` extracts a complete vector of user logins that made IDs for an observation
* `unnest()` converts the nested list structure into a flattened table
  - i.e., One row per observation x ID-er
* Result: a tidy table (`obs_tbl`) with columns:
  - obs_id
  - user_login

2. **Identify observations with trusted users**

```{r}
trusted_identifications <- obs_tbl %>%
  filter(user_login %in% trusted_users$user_login) %>%
  distinct(obs_id) %>%
  pull()

message(paste("There are", length(trusted_identifications), "iNat observations identified by trusted users.", sep = " " ))
```

4. **Subset the original dataset...**

```{r}
trusted_observations <- obs_api %>%
  filter(id %in% trusted_identifications)
```

```{r}
# Sanity check -- this should = TRUE
setequal(trusted_identifications, trusted_observations$id)
```

## Create a Map

`{rinat}` includes a mapping function that returns a ggplot map object, but you may choose to create your own with other tools.

```{r}
map <- inat_map(filtered_obs, plot = FALSE)

map + 
  annotation_borders("state") + 
  theme_classic()
```

That map sucks. It might depend on what your input is, but I'd like to make a better one...

```{r}
#| echo: true
#| output: false
library(sf)
library(maps)
```

0.  **Define target state**

    ```{r}
    target_state <- "alabama"
    ```

    *This is done purely for the easy reuse of code.*

1.  **Fetch target state boundary**

    ```{r}
    #| label: state-boundary
    state_boundary <- st_as_sf(maps::map("state", 
                                   regions = target_state, 
                                   fill = TRUE, plot = FALSE)) |>
                      st_set_crs(4326)
    ```


2.  **Convert inat data into an sf spatial object**

    ```{r}
    #| label: create-inat-sf
    inat_sf <- st_as_sf(filtered_obs, coords = c("longitude", "latitude"), crs = 4326)
    ```

3.  **Filter points to only keep the ones inside the boundary**

    ```{r}
    state_filtered_obs <- st_intersection(inat_sf, state_boundary)
    
    # Optional sanity check
    nrow(state_filtered_obs)
    nrow(filtered_obs)
    ```

4.  **Plot it!**

    ```{r}
    ggplot() +
      geom_sf(data = state_boundary, fill = "gray95", color = "black") +
      geom_sf(data = state_filtered_obs, color = "forestgreen", alpha = 0.8) + 
      theme_void()
    ```

### Add counties {#add-counties}

1.  Fetch county boundaries, filtered by target state

    ```{r}
    #| warning: false
    counties <- st_as_sf(maps::map("county", fill = TRUE, plot = FALSE)) %>%
      filter(grepl(target_state, ID)) |>
      st_set_crs(4326)
    ```

    :::{.callout-note}
    We already have the state boundary from [above](@state-boundary), so there's no need to generate it again.
    :::
    
2.  Plot it!

    ```{r}
    ggplot() +
      # Layer 1: County polygons with light borders
      geom_sf(data = counties, fill = "gray98", color = "gray80", size = 0.3) +
      
      # Layer 2: State outline (thick)
      geom_sf(data = state_boundary, fill = NA, color = "black", size = 0.8) +
      
      # Layer 3: Occurrence points 
      geom_sf(data = state_filtered_obs, color = "forestgreen", alpha = 0.5, size = 1.5) + 
      
      # Clean up background
      theme_void()
    ```

## Dealing with Uncertainty

You should be aware that the coordinates of iNat observations may be obscured. This can be determined on a by-record basis from the `coordinates_obscured` field which reflects the user's flagging (i.e., `geoprivacy`) and the iNat system designation (i.e., `taxon_geoprivacy`).

The "uncertainty" is recorded in `positional_accuracy` and `public_positional_accuracy`. 

::: {.callout-important}
To get the accurate coordinates, you would need to have the API return `private_latitude` and `private_longitude`. This requires establishing authentication AND being an authorized user (e.g., the observer, a "trusted user" of the observer's account, a project curator).

`**{rinat}**` does *not* support authentication.
:::

### Getting Accurate Coordinates

As mentioned before, you need to be an authorized user (e.g., the observer, a "trusted user" of the observer's account, a project curator).


#### Generate iNaturalist API Token {#generate-inat-token}

::: {.callout-note}
The steps detailed here generate a temporary (24-hour) token. I can work on an automation for it, but it requires the "app" to be registered with iNaturalist first. It won't be much of a hassle to do that, but it'll a little more time to pull the info for the application together.
:::

Go to [inaturalist.org/users/api_token](https://www.inaturalist.org/users/api_token). Then copy the complete string starting with `ey...` to your clipboard.

To avoid hardcoding the token directly in a script where it could be accidentally shared, you should save it as an environment variable. Open your user environment file...

```{r}
#| label: edit-r-environment
#| eval: false
if (!require("usethis")) install.packages("usethis")

usethis::edit_r_environ()
```

A text file called `.Renviron` will open in the editor (assuming you're using RStudio or another IDE). Add your token on a new line like this...

`INAT_TOKEN="paste_your_copied_token_here"`

Restart the R session for the changes to take effect.

```{r}
#| label: restart-r
#| eval: false
rstudioapi::restartSession()
```

::: {.callout-warning}
If you use Git/ GitHub, make sure your `.gitignore` file includes `.Renviron` so your token never leaves your computa!
:::

#### Fetch Obscured Data with `httr2` {#fetch-private-httr2}

Once your token is generated (see [Generate iNaturalist API Token](#generate-inat-token)), you can use use `Sys.getenv()` to query with the token securely within a script. 

See [iNaturalist v2 API](https://api.inaturalist.org/v2/docs/)

```{r}
library(httr2)
library(jsonlite)

# 1. Load token from .Renvir
my_jwt <- Sys.getenv("INAT_TOKEN")

# 2. Construct request targeting iNat v2 API
# -- Note: v2 requires you explicitly state what fields to return
req <- request("https://api.inaturalist.org/v2/observations") |>
  req_url_query(
    user_id = "emmerson_cray",
    geoprivacy = "obscured",
    fields = "all"
  ) |>
  # Pass the token as a Bearer string in the Authorization header
  req_headers(Authorization = my_jwt)

# 3. Execute the request
resp <- req_perform(req)
data <- resp_body_json(resp, simplifyVector = TRUE)

# 4. View results with real, hidden spatial locations intact
my_obs_df <- data$results
```

:::{.callout-note}
You can specify `fields = "all"` to return every single option, or be selective in the ones you want (e.g., `fields = "species_guess,observed_on,location,private_location,..."`)
:::

Here are all the fields the call above will return:

```{r}
#| echo: true
names(my_obs_df)
```

If you aren't being returned private fields, it's probably because your token is invalid. See [Troubleshoot Your Token](#troubleshoot-token) for help.

#### Fetch *Specific* Obscured Data with httr2

Since I won't have "trusted" access to the obscured coordinates from `filtered_obs`, I'll send an **unauthenticated** request to retrieve the observations by my user `emmerson_cray`...

```{r}
req_unauthenticated <- request("https://api.inaturalist.org/v2/observations") |>
  req_url_query(
    user_id = "emmerson_cray",
    fields = "all"
  )

resp_unauthenticated <- req_perform(req_unauthenticated)
data_unauthenticated <- resp_body_json(resp_unauthenticated, simplifyVector = TRUE)

my_obs_unauthenticated <- data_unauthenticated$results
```

And to compare the difference, I'll send an **authenticated** request to retrieve the observation by my user `emmerson_cray`...

```{r}
my_jwt <- Sys.getenv("INAT_TOKEN")

req_authenticated <- request("https://api.inaturalist.org/v2/observations") |>
  req_url_query(
    user_id = "emmerson_cray",
    fields = "all"
  ) |>
  req_headers(Authorization = my_jwt)

resp_authenticated <- req_perform(req_authenticated)
data_authenticated <- resp_body_json(resp_authenticated, simplifyVector = TRUE)

my_obs_authenticated <- data_authenticated$results
```


The API silently removes the columns that require some sort of authentication. We can tell by seeing the difference in the data frames...

```{r}
# Unauthenticated
ncol(my_obs_unauthenticated)

# Authenticated
ncol(my_obs_authenticated)
```

The exact columns that require authentication *(for this type of request)* are...

```{r}
library(janitor)
compare_df_cols(my_obs_unauthenticated, my_obs_authenticated) %>%
  filter(is.na(my_obs_unauthenticated)) |>
  select(column_name) |>
  gt()
```

Now that is explained, you can use this info to search for your target IDs of interest. 

1.  **Gather target IDs and combine into a single comma separated string**

    ```{r}
    # Extract the observation IDs from unauthenticated retrieval
    target_ids <- my_obs_unauthenticated %>%
      filter(geoprivacy == "obscured") %>%
      pull(id)
    
    # Combine into a single string
    id_string <- paste(target_ids, collapse = ",")
    ```

2.  **Query IDs using httr2**
```{r}
library(httr2)
library(jsonlite)

# 1. Load token from .Renvir
my_jwt <- Sys.getenv("INAT_TOKEN")

# 2. Construct request targeting iNat v2 API with target IDs
# -- Note: v2 requires you explicitly state what fields to return
req_target <- request("https://api.inaturalist.org/v2/observations") |>
  req_url_query(
    id = id_string,
    fields = "all"
  ) |>
  # Pass the token as a Bearer string in the Authorization header
  req_headers(Authorization = my_jwt)

# 3. Execute the request
resp_target <- req_perform(req_target)
data_target <- resp_body_json(resp_target, simplifyVector = TRUE)

# 4. View results with real, hidden spatial locations intact
target_obs <- data_target$results
```

```{r}
# Sanity check (should match ncol for authenticated)
ncol(target_obs) == ncol(my_obs_authenticated)
```

```{r}
target_obs %>%
  select(geoprivacy, location, private_location)
```

Then, if you wanted to parse these into lat/ long, you can do this easily by...
```{r}
target_obs %>%
  select(geoprivacy, location, private_location) %>%
  # Separate 'location'
  separate_wider_delim(
    cols = location,
    delim = ",",
    names = c("latitude", "longitude")
  ) %>%
    separate_wider_delim(
    cols = private_location,
    delim = ",",
    names = c("private_latitude", "private_longitude")
  )
```

#### Troubleshoot Your Token {#troubleshoot-token}

Is the authentication being used in the API request? 
```{r}
resp_headers(resp)[["X-User-Id"]]
resp$headers$`X-Cache`
```

If `X-User-Id` comes back as `NULL` and `X-Cache` is `"Miss"`, the API processed the request as a anonymous, unauthenticated public user. The API will silently filter out private fields without giving you an error.

```{r}
# Split the JWT to look at its payload
token_parts <- strsplit(Sys.getenv("INAT_TOKEN"), "\\.")[[1]]

if (length(token_parts) == 3) {
  # Decode the middle payload piece
  raw_payload <- rawToChar(jsonlite::base64_dec(token_parts[2]))
  payload <- jsonlite::fromJSON(raw_payload)
  
  # Print the expiration status
  expiration_time <- as.POSIXct(payload$exp, origin="1970-01-01", tz="UTC")
  print(paste("Token User ID:", payload$user_id))
  print(paste("Token expires at (UTC):", expiration_time))
  print(paste("Is token expired?", expiration_time < Sys.time()))
} else {
  print("The INAT_TOKEN string is not a valid JWT format.")
}

```

### Chloropleth Map 

For this example, I'll extend what we've already got from when we [added counties](#add-counties), but this could be any case where you've got polygons.

1.  Count points inside each county polygon

    ```{r}
    counts <- st_join(counties, inat_sf) %>%
      group_by(ID) %>%
      summarize(num_points = sum(!is.na(id)))
    ```

    ::: {.callout-note}
    This uses [`inat_sf`](@create-inat-sf), not any filtered version. 
    :::

2.  Plot counties shaded by data density

    ```{r}
    ggplot(data = counts) + 
      geom_sf(aes(fill = num_points), color = "gray50") +
      scale_fill_viridis_c(option = "cividis", name = "Observations") + 
      theme_void()
    ```

### Bonus! Thinking of accessibility...

While `scale_fill_viridis_c(option = "cividis")` is arguable the safest choice for color accessibility, there is a neat R package, `{scico}` that uses the [Crameri scientific color maps](https://www.fabiocrameri.ch/colourmaps/) that are colorblind-safe.

```{r}
#| eval: true
#| echo: true
# install.packages("scico")
library(scico)
```

`{scico}` has a lot of color palettes. YOu can see them all on the [Color Maps website](https://www.fabiocrameri.ch/colourmaps/) or call `{r} scico_palette_show()`

```{r}
#| label: scico-base
#| include: true
#| echo: true
base_plot <- ggplot(data = counts) + 
  geom_sf(aes(fill = num_points), color = "gray50") +
  theme_void() +
  theme(
    plot.title = element_text(
      size = 18,
      face = "bold",
      hjust = 0.5,
      vjust = 0.5
    )
  )
```

```{r}
#| label: scico-batlow
#| fig-cap: "batlow"
#| echo: true
p_batlow <- base_plot + 
  scale_fill_scico(palette = "batlow", name = "Observations") +
  ggtitle("batlow")
```

You may also choose to reverse the palette by using `direction`. For a lot of scico palettes, this will make it go from light to dark.

```{r}
#| label: scico-batlow-reversed
#| echo: true
p_batlow_reversed <- base_plot + 
  scale_fill_scico(palette = "batlow", direction = -1, , name = "Observations") +
  ggtitle("batlow (reversed)")
```

If you want to start from light and densities don't go below 0, you can still use a sequential scale with a little finesse. 

1.  Extract a smooth sequential gradient and place "white" at the start (0).

    ```{r}
    my_palette <- c("white", scico(10, palette = "batlow", direction = -1))
    ```

2.  Use `scale_fill_gradientn()` 

    ```{r}
    #| label: scico-my-palette
    #| echo: true
    p_my_palette <- base_plot +
      scale_fill_gradientn(colors = my_palette, name = "Observations") +
      ggtitle("my_palette")
    ```

```{r}
#| label: scico-plots
#| echo: false
library(patchwork)

p_batlow + p_batlow_reversed + p_my_palette &
  # Force color bar to be horizontal
  guides(fill = guide_colorbar(direction = "horizontal", title.position = "top")) &
  theme(
    # panel.background = element_rect(color = "black", linewidth = 1, fill = NA),
    plot.margin = margin(15, 15, 15, 15),
    legend.position = "inside",
    legend.position.inside = c(0.7, 0.1)
  )
```

## Packages Used

```{r}
#| echo: false
#| message: false
if(!require("grateful")) install.packages("grateful"); library(grateful)
scan_packages("Trusting_iNat_Observations.qmd", cite.tidyverse = FALSE) |>
  gt() |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

